---
title: "Creating Long-Format Peak Tables with Optional CAMERA Annotations"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Creating Long-Format Peak Tables with Optional CAMERA Annotations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

The `tidyXCMS` package provides functions to work with XCMS metabolomics data in
a tidy, long-format structure. This vignette demonstrates how to use the
`XCMSnExp_CAMERA_peaklist_long()` function to create a comprehensive peak table
that integrates:

- Peak detection results from XCMS
- Feature grouping (correspondence) information
- CAMERA annotations (isotopes, adducts, pseudospectrum groups) - **optional**
- Sample metadata

The resulting long-format table has one row per feature per sample, making it
ideal for downstream analysis with tidyverse tools like `dplyr` and `ggplot2`.

**Note:** CAMERA annotations are optional. You can create a peak table without
CAMERA annotations by simply omitting the `xsAnnotate` parameter.

# Setup

```{r setup, message=FALSE}
library(tidyXCMS)
library(xcms)
library(CAMERA)
library(commonMZ)
library(BiocParallel)
library(dplyr)
library(ggplot2)
library(tidyr)
library(MsExperiment)
```

# Example Workflow

## Load Example Data

We'll use the `xmse` dataset from XCMS, which contains preprocessed LC-MS data from
wild-type and knockout samples. This dataset already has peak detection and feature
grouping completed.

```{r load_data, message=FALSE}
# Load example data from XCMS (already preprocessed)
xdata <- loadXcmsData("xmse")

# Fix file paths to use correct system-specific location
# Extract relative paths (e.g., "KO/ko15.CDF", "WT/wt15.CDF")
current_paths <- fileNames(xdata)
relative_paths <- sub(".*/faahKO/cdf/", "", current_paths)

# Get the correct base path for faahKO package on this system
cdf_path <- file.path(find.package("faahKO"), "cdf")

# Reconstruct full paths
new_paths <- file.path(cdf_path, relative_paths)

# Update the file paths in the xdata object
fileNames(xdata) <- new_paths

# Check the sample information
sampleData(xdata)

# Check results
cat("Dataset contains", nrow(chromPeaks(xdata)), "peaks across",
    length(unique(chromPeaks(xdata)[, "sample"])), "samples,",
    "grouped into", nrow(featureDefinitions(xdata)), "features\n")
```

## Create Long-Format Peak Table (Without CAMERA)

You can create a long-format peak table directly from the XCMS results without using CAMERA annotations. This is useful when you want to quickly explore your data or when CAMERA annotations are not needed for your analysis.

```{r create_table_no_camera}
# Create peak table without CAMERA
peak_table_no_camera <- XCMSnExp_CAMERA_peaklist_long(xdata)

# Check the structure
dim(peak_table_no_camera)

# View first few rows
head(peak_table_no_camera)

# Check column names - note that isotopes, adduct, and pcgroup are NOT present
colnames(peak_table_no_camera)
```

The resulting table contains all feature and peak information. The CAMERA annotation columns (`isotopes`, `adduct`, `pcgroup`) are **not included** when no CAMERA annotations are provided, keeping the output cleaner and more memory-efficient.

## CAMERA Annotation (Optional)

Use CAMERA to annotate isotopes, adducts, and group features into pseudospectra.

**Note:** CAMERA requires an `xcmsSet` object. Since our data is an `XcmsExperiment`, we first convert it to `XCMSnExp` and then to `xcmsSet`.

### Creating a CAMERA object

In this first step we create a CAMERA object from the XCMS object. We specify the polarity to help CAMERA determine appropriate adducts and fragments.

```{r camera_setup, message=FALSE}
# Convert XcmsExperiment to xcmsSet for CAMERA (required for compatibility)
# Two-step conversion: XcmsExperiment -> XCMSnExp -> xcmsSet
xset <- xdata %>%
  as("XCMSnExp") %>%
  as("xcmsSet")

# Create xsAnnotate object with polarity
xs <- xsAnnotate(xset, polarity = "positive")
```

### Grouping coeluting peaks

The first step to grouping features is to group co-eluting peaks. This is a naïve approach that we will refine later.

```{r camera_groupfwhm, message=FALSE}
# Group peaks by retention time
xs <- groupFWHM(xs, perfwhm = 0.1, intval = "into", sigma = 6)
```

### Grouping based on correlation

Now we group the features based on correlations. This looks at each group from the previous step and splits them into separate groups for peaks that correlate with each other.

**Note:** For this small example dataset, don't use correlation across samples as it is unreliable with this few samples.

```{r camera_groupcorr, message=FALSE}
# Group by correlation
xs <- groupCorr(xs, calcIso = FALSE, calcCiS = TRUE, calcCaS = FALSE, cor_eic_th = 0.7, pval = 1E-6)
```

### Isotope annotation

This annotates peaks that are possible isotopes based on m/z difference and intensity patterns.

```{r camera_isotopes, message=FALSE}
# Find isotopes
xs <- findIsotopes(xs, ppm = 10, mzabs = 0.01, intval = "into", maxcharge = 2)
```

### Annotation of adducts and fragments

Now we try to annotate adducts and fragments based on expected mass differences. We'll use the `commonMZ` package to generate a comprehensive list of adduct and fragment rules for CAMERA.

First, we get the adduct/fragment rules for positive mode:

```{r camera_rules_filter, message=FALSE}
# warn_clash=TRUE will warn about indistinguishable annotations (e.g., [M+NH4]+ vs -NH3)
# Remove [M+NH4]+ if it's less common in your data

rules_pos <- MZ_CAMERA(mode = "pos", warn_clash = TRUE, clash_ppm = 5) %>%
  filter(name != "[M+NH4]+") %>%
  as.data.frame()
```

Now we can annotate adducts using these rules:

```{r camera_adducts, message=FALSE}
# Find adducts using the rules from commonMZ
xs <- findAdducts(xs, ppm = 500, mzabs = 0.2, multiplier = 4, polarity = "positive", rules = rules_pos)
```

## Create Long-Format Peak Table

Now we can create our comprehensive long-format peak table! Note that we use the original `xdata` (XCMSnExp) object along with the CAMERA annotations from `xs`.

```{r create_table}
peak_table <- XCMSnExp_CAMERA_peaklist_long(xdata, xs)

# Check the structure
dim(peak_table)
colnames(peak_table)
```

# Understanding the Output

The resulting tibble contains comprehensive information about each feature in each sample. Below is a complete description of all columns:

## Column Reference Table

The table has been organized into logical groups for easier understanding:

| Column | Content |
|--------|---------|
| **Sample information** |
| `filepath` | Path to the raw data file |
| `filename` | The filename without path |
| `fromFile` | The file number (the order files were supplied in) |
| *Plus any columns from `sampleData()`* | Sample metadata columns (e.g., sample_name, sample_group, sample_type) |
| **Feature identifiers** |
| `feature_id` | The index of the feature after grouping across samples |
| `peakidx` | The index of the peak before grouping across samples |
| **Feature-level m/z statistics (across all samples)** |
| `f_mzmed` | The median m/z found for that feature **across samples** |
| `f_mzmin` | The minimum m/z found for that feature across samples |
| `f_mzmax` | The maximum m/z found for that feature across samples |
| **Feature-level retention time statistics (across all samples)** |
| `f_rtmed` | The median retention time found for that feature across samples |
| `f_rtmin` | The minimum retention time found for that feature across samples |
| `f_rtmax` | The maximum retention time found for that feature across samples |
| **CAMERA annotations** |
| `isotopes` | The isotope annotation from CAMERA (e.g., "[M]+", "[M+1]+", "[M+2]+") |
| `adduct` | The adduct annotation from CAMERA (e.g., "[M+H]+", "[M+Na]+", "[M+NH4]+") |
| `pcgroup` | The feature grouping index from CAMERA - features with same ID likely come from the same compound |
| **Peak-level m/z measurements (in that specific sample)** |
| `mz` | The median m/z found for that feature in **that sample** |
| `mzmin` | The minimum m/z found for that feature in that sample |
| `mzmax` | The maximum m/z found for that feature in that sample |
| **Peak-level retention time measurements (in that specific sample)** |
| `rt` | The retention time found for that feature in that sample |
| `rtmin` | The minimum retention time found for that feature in that sample |
| `rtmax` | The maximum retention time found for that feature in that sample |
| **Peak intensity measurements** |
| `into` | The area under the peak |
| `intb` | The area under the peak after baseline removal |
| `maxo` | The maximum intensity (i.e., height) of the peak |
| `sn` | The signal to noise ratio of that peak |
| **Gaussian peak fitting parameters** |
| `egauss` | RMSE of Gaussian fit |
| `mu` | Gaussian parameter mu (center of the Gaussian; unit is scan number) |
| `sigma` | Gaussian parameter sigma |
| `h` | Gaussian parameter h (height of the Gaussian peak) |
| **CentWave algorithm parameters** |
| `f` | Region number of m/z ROI where the peak was localized |
| `dppm` | m/z deviation of mass trace across scans in ppm |
| `scale` | Scale on which the peak was localized |
| `scpos` | Center of peak position found by wavelet analysis |
| `scmin` | Left peak limit found by wavelet analysis (scan number) |
| `scmax` | Right peak limit found by wavelet analysis (scan number) |
| **Additional columns** |
| `ms_level` | MS level (e.g., MS1, MS2) |
| `is_filled` | Was the intensity found by gap filling (TRUE) or peak picking (FALSE) |

**Important notes:**

- **Each row represents one feature in one sample**
- If a feature was not detected in a sample, peak-level columns (mz, rt, into, etc.) will be `NA`
- Feature-level statistics (f_mzmed, f_rtmed, etc.) are always present and represent values across all samples
- CAMERA annotations apply at the feature level and are the same across all samples for a given feature
- The distinction between "feature-level" and "peak-level" is important:
  - **Feature-level** (prefix `f_`): Statistics aggregated across all samples
  - **Peak-level** (no prefix): Values for the specific peak in that specific sample

## Feature-Level Information

These columns describe each feature (grouped peak):

```{r feature_info}
peak_table %>%
  select(feature_id, f_mzmed, f_rtmed) %>%
  distinct() %>%
  head()
```

## CAMERA Annotations

- `isotopes`: Isotope annotation (e.g., "[M]+", "[M+1]+")
- `adduct`: Adduct annotation (e.g., "[M+H]+", "[M+Na]+")
- `pcgroup`: Pseudospectrum correlation group ID

```{r camera_info}
# View features with adduct annotations
peak_table %>%
  filter(adduct!="") %>%
  select(feature_id, f_mzmed, f_rtmed, isotopes, adduct, pcgroup) %>%
  distinct(feature_id, .keep_all = TRUE) %>%
  head()
```

## Peak-Level Information

For each feature in each sample:

- `mz`, `rt`: Detected peak position
- `into`: Integrated peak intensity
- `intb`: Baseline-corrected intensity
- `maxo`: Maximum intensity
- `sn`: Signal-to-noise ratio

```{r peak_info}
# View detected peaks (non-NA intensities)
peak_table %>%
  filter(!is.na(into)) %>%
  select(feature_id, filename, mz, rt, into, sn) %>%
  head()
```

## Sample Information

- `filename`, `filepath`: Sample file information
- `fromFile`: Sample index
- Plus any columns from `sampleData()` (sample metadata)

```{r sample_info}
peak_table %>%
  select(feature_id, filename, sample_name, sample_group, sample_type) %>%
  head()
```

## Missing Values

Features not detected in a sample have `NA` for peak-level columns:

```{r missing_values}
# Count detected features per sample
peak_table %>%
  group_by(filename) %>%
  summarise(
    n_features_detected = sum(!is.na(into)),
    n_features_total = n()
  )
```

# Downstream Analysis Examples

## Visualize Feature Intensities

```{r plot_intensities, fig.width=7, fig.height=4, fig.alt="Boxplot showing the distribution of log10-transformed feature intensities across sample groups. The plot displays the median, quartiles, and outliers for each group."}
# Plot intensity distribution by sample group
peak_table %>%
  filter(feature_id == 10) %>%
  ggplot(aes(x = sample_group, y = log10(into), fill = sample_group)) +
  geom_boxplot() +
  labs(
    title = "Feature Intensity Distribution",
    x = "Sample Group",
    y = "log10(Intensity)"
  ) +
  theme_minimal()
```

## Identify Features with Adduct Annotations

```{r adduct_summary}
# Count features by adduct type
adduct_counts <- peak_table %>%
  filter(adduct!="") %>%
  separate_rows(adduct, sep = "\\s(?=\\[M)") %>% 
  distinct(feature_id, adduct) %>%
  mutate(adduct = gsub("^(\\[.*\\].*\\+).*","\\1",adduct)) %>% 
  
  count(adduct, sort = TRUE)

adduct_counts
```

### Visualize Adduct Distribution

We can create a pie chart showing the frequency of different adducts and fragments found by CAMERA:

```{r adduct_pie, fig.width=7, fig.height=7, fig.alt="Pie chart showing the distribution of adduct and fragment annotations found by CAMERA across all features."}
# Create pie chart of adduct frequencies
pie(adduct_counts$n, labels = adduct_counts$adduct, main = "Distribution of Adduct Annotations")
```

## Find Features Present in All Samples

```{r complete_features}
# Features detected in all samples
complete_features <- peak_table %>%
  group_by(feature_id) %>%
  summarise(
    n_detected = sum(!is.na(into)),
    n_samples = n(),
    .groups = "drop"
  ) %>%
  filter(n_detected == n_samples)

cat("Found", nrow(complete_features), "features detected in all samples\n")
```

## Coefficient of Variation Analysis

```{r cv_analysis}
# Calculate CV for each feature
feature_cv <- peak_table %>%
  filter(!is.na(into)) %>%
  group_by(feature_id, f_mzmed, f_rtmed) %>%
  summarise(
    mean_intensity = mean(into),
    sd_intensity = sd(into),
    cv = sd_intensity / mean_intensity * 100,
    .groups = "drop"
  ) %>%
  arrange(cv)

# Show most stable features
head(feature_cv)
```

## Extract Specific Features for Further Analysis

```{r extract_features}
# Get a specific feature across all samples
feature_123 <- peak_table %>%
  filter(feature_id == 123) %>%
  select(feature_id, filename, sample_group, mz, rt, into)

head(feature_123)
```

# Session Info

```{r session_info}
sessionInfo()
```
