---
title: "Creating Long-Format Peak Tables with Optional CAMERA Annotations"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Creating Long-Format Peak Tables with Optional CAMERA Annotations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

The `tidyXCMS` package provides functions to work with XCMS metabolomics data in
a tidy, long-format structure. This vignette demonstrates how to use the
`XCMSnExp_CAMERA_peaklist_long()` function to create a comprehensive peak table
that integrates:

- Peak detection results from XCMS
- Feature grouping (correspondence) information
- CAMERA annotations (isotopes, adducts, pseudospectrum groups) - **optional**
- Sample metadata

The resulting long-format table has one row per feature per sample, making it
ideal for downstream analysis with tidyverse tools like `dplyr` and `ggplot2`.

**Note:** CAMERA annotations are optional. You can create a peak table without
CAMERA annotations by simply omitting the `xsAnnotate` parameter. This is useful
for quick data exploration or when CAMERA annotations are not required for your
analysis.

# Setup

```{r setup, message=FALSE}
library(tidyXCMS)
library(xcms)
library(CAMERA)
library(BiocParallel)
library(dplyr)
library(ggplot2)
```

# Example Workflow

## Load Example Data

We'll use the `faahko_sub` dataset from XCMS, which contains LC-MS data from
wild-type and knockout samples.

```{r load_data, message=FALSE}
# Load example data from XCMS
faahko_sub <- loadXcmsData("faahko_sub")

# Check the files
fileNames(faahko_sub)
```

## Add Sample Metadata

Let's add some sample information that will be included in our final table.

```{r add_metadata}
library(Biobase)

pd <- data.frame(
  sample_name = basename(fileNames(faahko_sub)),
  sample_group = c("WT", "KO", "WT")[1:length(fileNames(faahko_sub))],
  injection_order = 1:length(fileNames(faahko_sub)),
  row.names = basename(fileNames(faahko_sub))
)

pData(faahko_sub) <- pd
```

## Peak Detection

Perform chromatographic peak detection using the CentWave algorithm.

```{r peak_detection, message=FALSE}
# Define peak detection parameters
cwp <- CentWaveParam(
  peakwidth = c(20, 80),
  noise = 5000,
  ppm = 25
)

# Find peaks (using SerialParam for reproducibility)
xdata <- findChromPeaks(faahko_sub, param = cwp, BPPARAM = SerialParam())

# Check results
cat("Detected", nrow(chromPeaks(xdata)), "peaks across",
    length(fileNames(xdata)), "samples\n")
```

## Feature Grouping (Correspondence)

Group peaks across samples into features (aligned peaks).

```{r grouping, message=FALSE}
# Define grouping parameters
pdp <- PeakDensityParam(
  sampleGroups = pData(xdata)$sample_group,
  minFraction = 0.5,
  bw = 30
)

# Group peaks
xdata <- groupChromPeaks(xdata, param = pdp)

# Check results
cat("Grouped peaks into", nrow(featureDefinitions(xdata)), "features\n")
```

## Create Long-Format Peak Table (Without CAMERA)

You can create a long-format peak table directly from the XCMS results without using CAMERA annotations. This is useful when you want to quickly explore your data or when CAMERA annotations are not needed for your analysis.

```{r create_table_no_camera}
# Create peak table without CAMERA
peak_table_no_camera <- XCMSnExp_CAMERA_peaklist_long(xdata)

# Check the structure
dim(peak_table_no_camera)

# View first few rows
head(peak_table_no_camera)

# Check column names - note that isotopes, adduct, and pcgroup are not present
colnames(peak_table_no_camera)
```

The resulting table contains all feature and peak information, but the CAMERA annotation columns (`isotopes`, `adduct`, `pcgroup`) are not included.

## CAMERA Annotation (Optional)

Use CAMERA to annotate isotopes, adducts, and group features into pseudospectra.

**Note:** CAMERA requires an `xcmsSet` object, so we first convert the `XCMSnExp` object using `as("xcmsSet")`.

### Creating a CAMERA object

In this first step we create a CAMERA object from the XCMS object. We specify the polarity to help CAMERA determine appropriate adducts and fragments.

```{r camera_setup, message=FALSE}
# Convert to xcmsSet for CAMERA (required for compatibility)
xset <- as(xdata, "xcmsSet")

# Create xsAnnotate object with polarity
xs <- xsAnnotate(xset, polarity = "positive")
```

### Grouping coeluting peaks

The first step to grouping features is to group co-eluting peaks. This is a naïve approach that we will refine later.

```{r camera_groupfwhm, message=FALSE}
# Group peaks by retention time
xs <- groupFWHM(xs, perfwhm = 0.1, intval = "into", sigma = 6)
```

### Grouping based on correlation

Now we group the features based on correlations. This looks at each group from the previous step and splits them into separate groups for peaks that correlate with each other.

**Note:** For this small example dataset, we use a simplified approach. In real workflows with more samples, you would use stricter parameters and enable correlation across samples (`calcCaS = TRUE`).

```{r camera_groupcorr, message=FALSE}
# Group by correlation
# For small datasets, we skip correlation analysis to avoid errors
# In production with adequate samples, use:
# xs <- groupCorr(xs, calcCaS = TRUE, cor_eic_th = 0.7, pval = 0.05)

# For this example, we skip the correlation step and proceed directly
# xs <- groupCorr(xs, calcIso = FALSE, calcCiS = FALSE, calcCaS = FALSE)
```

### Isotope annotation

This annotates peaks that are possible isotopes based on m/z difference and intensity patterns.

```{r camera_isotopes, message=FALSE}
# Find isotopes
xs <- findIsotopes(xs, ppm = 10, mzabs = 0.01, intval = "into", maxcharge = 2)
```

### Annotation of adducts and fragments

Now we try to annotate adducts and fragments based on expected mass differences.

```{r camera_adducts, message=FALSE}
# Find adducts
xs <- findAdducts(xs, ppm = 10, mzabs = 0.01, multiplier = 4, polarity = "positive")
```

## Create Long-Format Peak Table

Now we can create our comprehensive long-format peak table! Note that we use the original `xdata` (XCMSnExp) object along with the CAMERA annotations from `xs`.

```{r create_table}
peak_table <- XCMSnExp_CAMERA_peaklist_long(xdata, xs)

# Check the structure
dim(peak_table)
colnames(peak_table)
```

# Understanding the Output

The resulting tibble contains comprehensive information about each feature in each sample. Below is a complete description of all columns:

## Column Reference Table

The table has been organized into logical groups for easier understanding:

| Column | Content |
|--------|---------|
| **Sample information** |
| `filepath` | Path to the raw data file |
| `filename` | The filename without path |
| `fromFile` | The file number (the order files were supplied in) |
| *Plus any columns from `pData(XCMSnExp)`* | Sample metadata columns you added |
| **Feature identifiers** |
| `feature_id` | The index of the feature after grouping across samples |
| `peakidx` | The index of the peak before grouping across samples |
| **Feature-level m/z statistics (across all samples)** |
| `f_mzmed` | The median m/z found for that feature **across samples** |
| `f_mzmin` | The minimum m/z found for that feature across samples |
| `f_mzmax` | The maximum m/z found for that feature across samples |
| **Feature-level retention time statistics (across all samples)** |
| `f_rtmed` | The median retention time found for that feature across samples |
| `f_rtmin` | The minimum retention time found for that feature across samples |
| `f_rtmax` | The maximum retention time found for that feature across samples |
| **CAMERA annotations** |
| `isotopes` | The isotope annotation from CAMERA (e.g., "[M]+", "[M+1]+", "[M+2]+") |
| `adduct` | The adduct annotation from CAMERA (e.g., "[M+H]+", "[M+Na]+", "[M+NH4]+") |
| `pcgroup` | The feature grouping index from CAMERA - features with same ID likely come from the same compound |
| **Peak-level m/z measurements (in that specific sample)** |
| `mz` | The median m/z found for that feature in **that sample** |
| `mzmin` | The minimum m/z found for that feature in that sample |
| `mzmax` | The maximum m/z found for that feature in that sample |
| **Peak-level retention time measurements (in that specific sample)** |
| `rt` | The retention time found for that feature in that sample |
| `rtmin` | The minimum retention time found for that feature in that sample |
| `rtmax` | The maximum retention time found for that feature in that sample |
| **Peak intensity measurements** |
| `into` | The area under the peak |
| `intb` | The area under the peak after baseline removal |
| `maxo` | The maximum intensity (i.e., height) of the peak |
| `sn` | The signal to noise ratio of that peak |
| **Gaussian peak fitting parameters** |
| `egauss` | RMSE of Gaussian fit |
| `mu` | Gaussian parameter mu (center of the Gaussian; unit is scan number) |
| `sigma` | Gaussian parameter sigma |
| `h` | Gaussian parameter h (height of the Gaussian peak) |
| **CentWave algorithm parameters** |
| `f` | Region number of m/z ROI where the peak was localized |
| `dppm` | m/z deviation of mass trace across scans in ppm |
| `scale` | Scale on which the peak was localized |
| `scpos` | Center of peak position found by wavelet analysis |
| `scmin` | Left peak limit found by wavelet analysis (scan number) |
| `scmax` | Right peak limit found by wavelet analysis (scan number) |
| **Additional columns** |
| `ms_level` | MS level (e.g., MS1, MS2) |
| `is_filled` | Was the intensity found by gap filling (TRUE) or peak picking (FALSE) |

**Important notes:**

- **Each row represents one feature in one sample**
- If a feature was not detected in a sample, peak-level columns (mz, rt, into, etc.) will be `NA`
- Feature-level statistics (f_mzmed, f_rtmed, etc.) are always present and represent values across all samples
- CAMERA annotations apply at the feature level and are the same across all samples for a given feature
- The distinction between "feature-level" and "peak-level" is important:
  - **Feature-level** (prefix `f_`): Statistics aggregated across all samples
  - **Peak-level** (no prefix): Values for the specific peak in that specific sample

## Feature-Level Information

These columns describe each feature (grouped peak):

```{r feature_info}
peak_table %>%
  select(feature_id, f_mzmed, f_rtmed) %>%
  distinct() %>%
  head()
```

## CAMERA Annotations

- `isotopes`: Isotope annotation (e.g., "[M]+", "[M+1]+")
- `adduct`: Adduct annotation (e.g., "[M+H]+", "[M+Na]+")
- `pcgroup`: Pseudospectrum correlation group ID

```{r camera_info}
# View features with adduct annotations
peak_table %>%
  filter(!is.na(adduct)) %>%
  select(feature_id, f_mzmed, f_rtmed, isotopes, adduct, pcgroup) %>%
  distinct(feature_id, .keep_all = TRUE) %>%
  head()
```

## Peak-Level Information

For each feature in each sample:

- `mz`, `rt`: Detected peak position
- `into`: Integrated peak intensity
- `intb`: Baseline-corrected intensity
- `maxo`: Maximum intensity
- `sn`: Signal-to-noise ratio

```{r peak_info}
# View detected peaks (non-NA intensities)
peak_table %>%
  filter(!is.na(into)) %>%
  select(feature_id, filename, mz, rt, into, sn) %>%
  head()
```

## Sample Information

- `filename`, `filepath`: Sample file information
- `fromFile`: Sample index
- Plus any columns from `pData()` (sample metadata)

```{r sample_info}
peak_table %>%
  select(feature_id, filename, sample_group, injection_order) %>%
  head()
```

## Missing Values

Features not detected in a sample have `NA` for peak-level columns:

```{r missing_values}
# Count detected features per sample
peak_table %>%
  group_by(filename) %>%
  summarise(
    n_features_detected = sum(!is.na(into)),
    n_features_total = n()
  )
```

# Downstream Analysis Examples

## Visualize Feature Intensities

```{r plot_intensities, fig.width=7, fig.height=4}
# Plot intensity distribution by sample group
peak_table %>%
  filter(!is.na(into)) %>%
  ggplot(aes(x = sample_group, y = log10(into), fill = sample_group)) +
  geom_boxplot() +
  labs(
    title = "Feature Intensity Distribution",
    x = "Sample Group",
    y = "log10(Intensity)"
  ) +
  theme_minimal()
```

## Identify Features with Adduct Annotations

```{r adduct_summary}
# Count features by adduct type
peak_table %>%
  filter(!is.na(adduct)) %>%
  distinct(feature_id, adduct) %>%
  count(adduct, sort = TRUE)
```

## Find Features Present in All Samples

```{r complete_features}
# Features detected in all samples
complete_features <- peak_table %>%
  group_by(feature_id) %>%
  summarise(
    n_detected = sum(!is.na(into)),
    n_samples = n(),
    .groups = "drop"
  ) %>%
  filter(n_detected == n_samples)

cat("Found", nrow(complete_features), "features detected in all samples\n")
```

## Coefficient of Variation Analysis

```{r cv_analysis}
# Calculate CV for each feature
feature_cv <- peak_table %>%
  filter(!is.na(into)) %>%
  group_by(feature_id, f_mzmed, f_rtmed) %>%
  summarise(
    mean_intensity = mean(into),
    sd_intensity = sd(into),
    cv = sd_intensity / mean_intensity * 100,
    .groups = "drop"
  ) %>%
  arrange(cv)

# Show most stable features
head(feature_cv)
```

## Extract Specific Features for Further Analysis

```{r extract_features}
# Get a specific feature across all samples
feature_123 <- peak_table %>%
  filter(feature_id == 123) %>%
  select(feature_id, filename, sample_group, mz, rt, into)

head(feature_123)
```

# Working with chromPeaks_classic()

The `chromPeaks_classic()` helper function extracts all chromatographic peaks
as a tibble, which can be useful for exploratory analysis:

```{r chrompeaks_classic}
# Extract all peaks
all_peaks <- chromPeaks_classic(xdata)

# Check structure
dim(all_peaks)
head(all_peaks)

# Analyze peak properties
all_peaks %>%
  group_by(sample) %>%
  summarise(
    n_peaks = n(),
    median_sn = median(sn),
    median_into = median(into)
  )
```

# Session Info

```{r session_info}
sessionInfo()
```
